Viscoelastic Kelvin-Helmholtz instability

### 2016-02-12 paper/paper.tex
updated paper with more new plots, improved ordering of some plots

%	paper.tex	Toby Searle
%	Last modified: Fri 12 Feb 11:52:33 2016
%	using jfm template

@@

% NOTE: A full address must be provided: department, university/institution, town/city, zipcode/postcode, country.
\affiliation{$^1$SUPA, School of Physics and Astronomy, University of Edinburgh, Mayfield Road,
Edinburgh, EH9 3JZ, UK\\[\affilskip]
}

@@

It was long thought that the elasticity of a viscoelastic fluid only served to
dampen instability. This view stems from early results showing that drag in a
turbulent Newtonian flow can be reduced by the addition of small amounts of
viscoelastic fluid \citep{Toms1977}. Although drag reduction has been studied
for a long time, studies on turbulence without inertia \citep{Larson1990,
Groisman2000} give reason to believe that there are purely elastic bulk
instabilities present in elastic fluids relevant to industry.  This purely
elastic turbulence is present at very low Reynolds number, where it is driven
by the elasticity of a polymeric flow rather than its inertia.

@@

\citet{Azaiez1994} do the linear stability of a free shear flow of the
Oldroyd-B, co-rotational Jefferys and Gisekus fluids. Find suppression of the
Newtonian Kelvin-Helmholtz instability. They do asymptotic long wave linear
stability such that the Reynold's number is small but finite and the
Weissenberg number is large with a constant elasticity. They also do a
numerical linear stability via an orthogonal shooting algorithm. In the
appendix of this paper Hinch \citep{Azaiez1994} takes the long wavelength limit
of their problem in a far more elegant way and exposes a stabilisation
mechanism based on elastic hoop stresses in the curved streamlines.

@@

We used an Oldroyd-B fluid with very low Reynolds number $\Rey \leq 1.0$ and a
small ratio between the solvent and total viscosities $\beta =
\frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid,
where the polymer contribution to the stress on the fluid is much larger  than
the solvent contribution. The Oldroyd-B Navier stokes and constitutive
equations are then: 
\begin{align} 
    Re \left[ \dt{\mathbf{v}} +
    \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta
    \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T}
    + \nabla{\mathbf{v}} 
\end{align} 
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$
as the total stress in the polymeric fluid. The FENE-P viscoelastic fluid has
the advantage of being similarly easy to analyse using our spectral method, but
also including a finite extensibility for the polymers. This leads to a stress
which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells
via a non-linear spring force:

@@

\begin{figure} 
    \centering 
    \includegraphics[width=0.3\textwidth]{{KH_diagram}}
    \caption{diagram of the system} 
    \label{fig:KH_diagram} 
\end{figure}

@@

\begin{figure} 
    \centering 
    \begin{subfigure}[b]{0.48\textwidth} 
	\centering
	\includegraphics[width=\textwidth]{{inf_purely_elastic}}
	\caption{}
	\label{fig:KH_inf_disp_height_vary_Wi} 
    \end{subfigure} ~
    \begin{subfigure}[b]{0.48\textwidth} 
	\centering
	\includegraphics[width=\textwidth]{{inf_dispersions_low_Re}} 
	\caption{}
	\label{fig:KH_inf_dispersions_vary_Wi} 
    \end{subfigure} 
    \caption{
	a) Free shear version of the instability. Plot of the maximum growth
	rate against Weissenberg number at $\Rey = 0.01$. b) Dispersion
	relations at various Weissenberg numbers to accompany the trend in
	figure a).
    } 
\end{figure}

@@

Now we use equations \ref{eq:KH_laminar_profile} and examine the unstretched
pseudospectral domain.  The instability remains present under variation of
$\Delta$ at low $\Delta$. At $\Delta =0.1$ the instability is still present for
$\Rey < 0.01$ far lower than the usual threshold of $\Rey \sim 1000$ for a
Newtonian Kelvin-Helmholtz instability (figure \ref{fig:KH_walls_vary_Re}).

@@

\begin{figure} 
    \centering 
    \includegraphics[width=\textwidth]{{KH_vary_delta}}
    \caption{
	Dispersion relations for various values of $\Delta$ at $\Rey = 0.1$,
	$\Wi = 100$. The sensitivity to the walls seems only to increase the
	strength of the instability. Even when the system is insensitive to the
	walls, at $\Delta= 0.0125$, there is still an instability present.
    } 
    \label{fig:KH_low_Re_walls_dependence} 
\end{figure}

\begin{figure} \centering
    \includegraphics[width=0.7\textwidth]{{high_Re_vary_delta}}
    \caption{
	Dispersion relations for various values of $\Delta$ and $\Rey = 1000$,
	$\Wi = 0.01$. When results are rescaled, the instability is
	found to be insensitive for $\Delta =0.1$.
    }
    \label{fig:KH_high_Re_walls_dependence}
\end{figure}

@@

\begin{figure}[h]
    \includegraphics[width=\textwidth]{{Visco_Newt_slice}.pdf}
    \caption{
	Streamfunction of the Newtonian Kelvin Helmholtz instability at $\Rey =
	1000$ compared with the purely elastic instability for $\beta = 0.1$,
	$\Rey = 0.0$, $\Wi = 5.0$, both at $\Delta =0.05$. The Newtonian Kelvin
	Helmholtz instability exponentially decays towards the channel walls
	whilst large vortices are present outside the walls in the purely
	elastic case.
    }
    \label{fig:KH_Visco_Newt_evectors}
\end{figure}

@@

\begin{figure} 
    \centering 
    \includegraphics[width=\textwidth]{{psi_high_delta}}
    \caption{ 
	Streamfunction of the eigenfunction of the instability at $\Delta =
	0.1$, $\Wi = 100$, $\Rey=0.1$. The magnitude is scaled by the xx stress
	in the centre of the channel.  
    } 
    \label{fig:KH_psi_high_delta}
\end{figure}

@@

\begin{figure} \centering
    \includegraphics[width=0.7\textwidth]{{vary_Re_delta_conv}}
    \caption{ The
	data in red corresponds to the free shear system and the green gives
	the $\Delta = 0.1$ data. By $\Rey = 1$ it is clear that the Reynold's
	number is sufficient such that the walled system behaves as though it
    were insensitive to the walls.  } \label{fig:KH_disp_height_vary_Re_delta_conv}
\end{figure}

@@

\begin{figure} 
    \centering
    \includegraphics[width=0.7\textwidth]{{growth_rate_sldim}} 
    \caption{ 
	Phase diagram showing the growth rate for various values of the
	Reynold's and Weissenberg numbers at $\beta=0.1$. Grey lines represent
	lines of constant elasticity. Black squares are data points for which
	the flow is linearly stable.  
    } 
    \label{fig:KH_growth_rate_phase_diagram} 
\end{figure}

@@

The finite extensibility version of the model gives good agreement with the
Oldroyd-B version. Shorter lengths bring about a stronger instability and do not
delay it to higher Weissenberg number. Otherwise similar dispersion relations
are observed with a similar dependence on the distance to the walls (figure
\ref{fig:KH_FENE_low_Re}).

\begin{figure} 
    \centering
    \includegraphics[width=0.7\textwidth]{{FENE_purely_elastic}} 
    \caption{ 
	The FENE-P model fluid at $\Rey = 0$  when $\Delta = 0.1$, ranging from
	$L=100$ (top curve), $L=500$ and the Oldroyd-B model in the bottom
	curve. Results are consistent with figure \ref{fig:KH_purely_elastic}.  
    }
	\label{fig:KH_FENE_low_Re} \end{figure}

We see the height of the instability decrease with increasing Weissenburg number, and this trend is stronger for smaller $\beta$ (figure \ref{fig:KH_drag_reduction}).

\begin{figure}
%   \showthe\columnwidth %use this to determine size of figures.
    \centering \includegraphics[width=0.7\textwidth]{{KH_high_Re_vary_Wi}}
    \caption{ 
	Peak growth rate against Weissenberg number at $\Rey = 1000$,
	$\Delta = 0.1$ The height of the dispersion relation is reduced for the
	Kelvin-Helmholtz instability as the Weissenberg number is increased.
	The grey dashed line is the height of the dispersion relation for the
	Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased
	stability of the base flow, this matches the observations made in \cite{Azaiez1994}} 
    \label{fig:KH_drag_reduction} 
\end{figure}

>>>>

%	paper.tex	Toby Searle
%	Last modified: Fri 12 Feb 17:42:28 2016
%	using jfm template

@@

% NOTE: A full address must be provided: department, university/institution, town/city, zipcode/postcode, country.
\affiliation{$^1$SUPA, School of Physics and Astronomy, University of Edinburgh, Mayfield Road,
Edinburgh, EH9 3JZ, UK\\
[\affilskip]
}

@@

The Kelvin-Helmholtz instability is integral to the transition to turbulence in
Newtonian fluids. By finding a purely elastic version of this instability, we
seek to probe a generic mechanism for the transition to purely elastic
turbulence, one that is presumably present in other contexts, such as the
oscillatory channel flow problem of \cite{Searle2016b}. To explore this
mechanism in the purely elastic regime we have constructed a simple model
system of a purely elastic flow which generates large jumps in the first normal
stress difference and so ought to lead to instability.

This flow has not been considered in the past because it does not occur in many
of the natural contexts where we see the Newtonian Kelvin-Helmholtz. For
example, the archetypical Kelvin-Helmholtz instability takes place for flow
over a backwards facing step. In this geometry, the fluids inertia leads to an
overshoot with vortex behind the step and a hyperbolic tangent streamwise flow
velocity. In Purely elastic fluids the flow is Stokesian, and so the fluid
velocity is much smoother with no overshoot.

The backwards facing step is not the flow we are thinking of when we examine
this instability, however. We are thinking of flows where elastic effects lead
to sharp changes in velocity and stress. Such jumps in first normal stress
difference are thought to lead to an instability see \citet{Morozov2007}, and
we hope to expose this as a generic mechanism by which a flow might become
unstable.

@@

\citet{Azaiez1994} do the linear stability of a free shear flow of the
Oldroyd-B, co-rotational Jefferys and Gisekus fluids. For large elasticity they
find suppression of the Newtonian Kelvin-Helmholtz instability. They do
asymptotic long wave linear stability such that the Reynold's number is small
but finite and the Weissenberg number is large with a constant elasticity. They
also do a numerical linear stability via an orthogonal shooting algorithm. In
the appendix of this paper Hinch \citep{Azaiez1994} takes the long wavelength
limit of their problem in a far more elegant way and exposes a stabilisation
mechanism based on elastic hoop stresses in the curved streamlines.

@@

We used an Oldroyd-B fluid with very low Reynolds number $\Rey \leq 1.0$ and a
small ratio between the solvent and total viscosities $\beta =
\frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid,
where the polymer contribution to the stress on the fluid is much larger  than
the solvent contribution. The Oldroyd-B Navier stokes and constitutive
equations are then: 
\begin{align} 
    Re \left[ \dt{\mathbf{v}} +
    \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta
    \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T}
    + \nabla{\mathbf{v}} 
\end{align} 
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$
as the total stress in the polymeric fluid. The FENE-CR viscoelastic fluid has
the advantage of being similarly easy to analyse using our spectral method, but
also including a finite extensibility for the polymers. This leads to a stress
which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells
via a non-linear spring force:

@@

\begin{figure}[h]
    \centering 
    \includegraphics[width=0.3\textwidth]{{KH_diagram}}
    \caption{Diagram of a shear layer} 
    \label{fig:KH_diagram} 
\end{figure}

@@

\begin{figure} 
    \centering 
    \begin{subfigure}[b]{0.48\textwidth} 
	\centering
	\includegraphics[width=\textwidth]{{inf_purely_elastic}}
	\caption{}
	\label{fig:KH_inf_disp_height_vary_Wi} 
    \end{subfigure} 
    ~
    \begin{subfigure}[b]{0.48\textwidth} 
	\centering
	\includegraphics[width=\textwidth]{{inf_dispersions_Re0.01}.pdf} 
	\caption{}
	\label{fig:KH_inf_dispersions_vary_Wi} 
    \end{subfigure} 
    \caption{
	a) Free shear version of the instability. Plot of the maximum growth
	rate against Weissenberg number at $\Rey = 0.01$. b) Dispersion
	relations at various Weissenberg numbers to accompany the trend in
	figure a).
    } 
\end{figure}

@@

Now we use equations \ref{eq:KH_laminar_profile} and examine the unstretched
pseudospectral domain. To validate the code, we examine the high Reynolds
number regime. We see the height of the instability decrease with increasing
Weissenburg number, and this trend is stronger for smaller $\beta$ (figure
\ref{fig:KH_drag_reduction}). This is the same behaviour reported in
\cite{Azaiez1994}.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{{high_Re_vary_delta}}
    \caption{
	Dispersion relations for various values of $\Delta$ and $\Rey = 1000$,
	$\Wi = 0.01$. When results are rescaled, the instability is
	found to be insensitive for $\Delta =0.1$.
    }
    \label{fig:KH_high_Re_walls_dependence}
\end{figure}

\begin{figure}[h]
%   \showthe\columnwidth %use this to determine size of figures.
    \centering \includegraphics[width=0.7\textwidth]{{KH_high_Re_vary_Wi}}
    \caption{ 
	Peak growth rate against Weissenberg number at $\Rey = 1000$,
	$\Delta = 0.1$ The height of the dispersion relation is reduced for the
	Kelvin-Helmholtz instability as the Weissenberg number is increased.
	The grey dashed line is the height of the dispersion relation for the
	Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased
	stability of the base flow, this matches the observations made in \cite{Azaiez1994}} 
    \label{fig:KH_drag_reduction} 
\end{figure}

The instability remains present under variation of $\Delta$ at low $\Delta$. At
$\Delta =0.1$ the instability is still present for $\Rey < 0.01$ far lower than
the usual threshold of $\Rey \sim 1000$ for a Newtonian Kelvin-Helmholtz
instability (figure \ref{fig:KH_walls_vary_Re}).

@@

\begin{figure}[h]
    \centering 
    \includegraphics[width=\textwidth]{{KH_vary_delta}}
    \caption{
	Dispersion relations for various values of $\Delta$ at $\Rey = 0.1$,
	$\Wi = 100$. The sensitivity to the walls seems only to increase the
	strength of the instability. Even when the system is insensitive to the
	walls, at $\Delta= 0.0125$, there is still an instability present.
    } 
    \label{fig:KH_low_Re_walls_dependence} 
\end{figure}

@@

\begin{figure}[h]
    \includegraphics[width=\textwidth]{{visco_Newt_slice}.pdf}
    \caption{
	Streamfunction of the Newtonian Kelvin Helmholtz instability at $\Rey =
	100$, $k=0.722$ compared with the purely elastic Oldroyd-B and FENE-CR
	instability for $\beta = 0.1$, $\Rey = 0.0$, $k=0.02$, $\Wi = 50.0$,
	both at $\Delta =0.05$. Newtonian eigenvector is in red, with the
	Oldroyd-B in green and the FENE model in dashed purple at $L=100$. The
	Newtonian Kelvin Helmholtz instability exponentially decays towards the
	channel walls whilst large vortices are present outside the shear layer
	in the purely elastic case.
    }
    \label{fig:KH_Visco_Newt_evectors}
\end{figure}

@@

\begin{figure}[h]
    \centering 
    \includegraphics[width=\textwidth]{{psi_high_delta}}
    \caption{ 
	Streamfunction of the eigenfunction of the instability at $\Delta =
	0.1$, $\Wi = 100$, $\Rey=0.1$. The magnitude is scaled by the xx stress
	in the centre of the channel.  
    } 
    \label{fig:KH_psi_high_delta}
\end{figure}

@@

\begin{figure}[h] 
    \centering
    \includegraphics[width=0.7\textwidth]{{vary_Re_delta_conv}}
    \caption{ The
	data in red corresponds to the free shear system and the green gives
	the $\Delta = 0.1$ data. By $\Rey = 1$ it is clear that the Reynold's
	number is sufficient such that the walled system behaves as though it
    were insensitive to the walls.  } \label{fig:KH_disp_height_vary_Re_delta_conv}
\end{figure}

@@

\begin{figure}[h] 
    \centering
    \includegraphics[width=0.7\textwidth]{{growth_rate_sldim}} 
    \caption{ 
	Phase diagram showing the growth rate for various values of the
	Reynold's and Weissenberg numbers at $\beta=0.1$. Grey lines represent
	lines of constant elasticity. Black squares are data points for which
	the flow is linearly stable.  
    } 
    \label{fig:KH_growth_rate_phase_diagram} 
\end{figure}

@@

The finite extensibility version of the model gives good agreement with the
Oldroyd-B version. Shorter lengths bring about a stronger instability and do
not delay it to higher Weissenberg number. Otherwise similar dispersion
relations are observed with a similar dependence on $\Delta$. (figure
\ref{fig:KH_FENE_low_Re}). The dependence of the instability on wavenumber is
not shown, but is exactly the same as that seen for the Oldroyd-B model.

A shorter length scale for the polymers results in an amplification of the
instability. This suggests that perhaps the instability mechanism involves the
shear stress on the polymers.

\begin{figure}[h] 
    \centering
    \includegraphics[width=0.7\textwidth]{{FENE_purely_elastic}} 
    \caption{ 
	The FENE-CR model fluid at $\Rey = 0$  when $\Delta = 0.1$, ranging from
	$L=100$ (top curve), $L=500$ and the Oldroyd-B model in the bottom
	curve. Results are consistent with figure \ref{fig:KH_purely_elastic}.  
    }
    \label{fig:KH_FENE_low_Re} 
\end{figure}
